Model-based pretreatment for regression with correlated variables.

### 2014-04-15 www/article13.tex
up
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/correg@537 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\section{Model to select decorrelated covariates}
\subsection{A classical problem: correlations in regression}\label{sectionOLS}
For a model defined by (\ref{regressionsimple}), we have the following Ordinary Least Squares (\textsc{OLS}) estimator:
	\begin{equation}
		\boldsymbol{\hat{\beta}}=\left(\boldsymbol{X}'\boldsymbol{X} \right) ^{-1}\boldsymbol{X}'\boldsymbol{Y}
	\end{equation}
	with variance matrix
	\begin{equation}
		\operatorname{Var}(\boldsymbol{\hat{\beta}})=\sigma_Y^2\left(\boldsymbol{X}'\boldsymbol{X} \right) ^{-1} \label{eqOLS}
	\end{equation}
	and without any bias.
	When correlations between covariates are strong, the matrix to invert is ill-conditioned and the variance increases, giving unstable and unusable estimator \cite{hoerl1970ridge}.
	Another problem is that matrix inversion requires $n\geq p$. 
\subsection{Our proposal: modelisation of the correlations}
Let now $\boldsymbol{X} \in \mathcal{R}^{n\times p}$ be a set of $p$ covariates.
In the following, we note $\boldsymbol{X^j}$ the $j^{th}$ column of $\boldsymbol{X}$.

We make the hypothesis that correlations in $\boldsymbol{X}$ are generated by a family of $p_r$ internal regressions between covariates with $I_r$ the set of indices of endogenous variables in $\boldsymbol{X}$ (explained ones). We note $\boldsymbol{X}_r=\boldsymbol{X}^{I_r}$ the sub-matrix of the redundant covariates (explained by others) and  $\boldsymbol{X}_f$ the remaining submatrix of free covariates (independent explaining covariates).
We focus here on the expression of $P(\boldsymbol{X}_r|\boldsymbol{X}_f)$.

We have $I_f=\{I_f^1,\dots,I_f^p \}$ the set of the sets of indices of exogenous covariates (explaining ones $=\boldsymbol{X}_f$) with $\forall j \notin I_r, I_f^j=\emptyset$. Then we have an explicit structure $S=(I_f,I_r,p_f,p_r)$ where $p_r=|I_r|$, $\boldsymbol{p}_f=(p_f^1,\dots,p_f^{p_r})$ is the vector of the number of covariates in each internal regression  and $p_f^j=|I_f^j|$. Here $|.|$ represents the cardinal of an ensemble.

The family of $p_r$ regressions within correlated covariates in $\boldsymbol{X}$ is noted:
	\begin{equation}
		\boldsymbol{X}_{r|\boldsymbol{X}_f,S} \textrm{ defined by }\forall j \in I_r: \boldsymbol{X}^j_{|\boldsymbol{X}_f,S}=\boldsymbol{X}_f\boldsymbol{\alpha}_j+\boldsymbol{\varepsilon}_j \textrm{ with } \boldsymbol{\varepsilon}_j \sim(\boldsymbol{0},\sigma^2_j\boldsymbol{I}_n) \label{SR}
	\end{equation}
		where $\boldsymbol{\alpha}_j \in \mathcal{R}^{(p-p_2)}$ are the vectors of the regression coefficients between the covariates (containing some zeros according to $I_f^j$).

Now we have an explicit structure of correlations and we know which covariates will cause ill-conditioning. The explicit form also gives the ability to understand phenomenon inside the dataset and to see why some covariates are not selected by methods like the LASSO when all $\boldsymbol{X}$ is provided (redundent variables).

We make the hypothesis of the uncrossing rule $I_f\cap I_r=\emptyset$ {\it i.e.} endogenous variables don't explain other covariates, thus $\boldsymbol{X}_r$ and $\boldsymbol{X}_f$ give a partition of $\boldsymbol{X}$, so $\boldsymbol{X}_f=\boldsymbol{X}\setminus \boldsymbol{X}_r$. This hypothesis ensures that $S$ contains no cycle and is straightforward readable (no need to order $I_r$). This hypothesis is not so restrictive because cyclic structures have no sense and any non-cyclic structure can be associated with a structure that verifies the uncrossing constraint by just replacing covariates in $\boldsymbol{X}_r$ by their expression in $S$ when on the right of a sub-regression.

@@

	  We make the choice to distinguish the response variable from the other endogenous variables (that are on the left of a regression). Thus we have one regression on our response variable ($P(\boldsymbol{Y}|\boldsymbol{X})$ and a system of sub-regressions (without the response variable: $P(\boldsymbol{X}_r|\boldsymbol{X}_f,S)$). Then we consider correlations between the explicative covariates of the main regression, not between the residuals. We see that the $S$ does not depend on $Y$ so it can be learnt independently, even with a larger dataset (if missing values in $Y$).

@@

\subsection{Strategy of use: pretreatment before classical estimation/selection methods}\label{interpretation}
As a pretreatment, it allows usage of any method in a second time to estimate $\boldsymbol{\beta}^*$, even with variable selection methods like LASSO or even stepwise \cite{seber2012linear}.

@@

	For the marginal regression model defined in (\ref{modexpl})
%	\begin{equation}
%		\boldsymbol{Y}_{|\boldsymbol{X}_f}= \boldsymbol{X}_f\boldsymbol{\beta}_f^*+ \boldsymbol{\varepsilon}_Y^*
%	\end{equation}			
%		So 
we have the \textsc{OLS} unbiased estimator of $\boldsymbol{\beta}^*$: 
		\begin{equation}
			\hat{\boldsymbol{\beta}}_f^* = (\boldsymbol{X}_f'\boldsymbol{X}_f)^{-1}\boldsymbol{X}_f'\boldsymbol{Y}  \textrm{ and }\boldsymbol{\hat\beta}_r^* = \boldsymbol{0}
		\end{equation}
		We see in (\ref{Trueexpl}) that it gives an unbiased estimation of $\boldsymbol{Y}$ and $\boldsymbol{\beta^*}$
		but in terms of $\boldsymbol{\beta}$ this estimator is biased:
		\begin{equation}
			\operatorname{E}[\hat{\boldsymbol{\beta}}_f^*|\boldsymbol{X}_f]=\boldsymbol{\beta}_f+\sum_{j \in I_r}\beta_{j}\boldsymbol{\alpha}_j \textrm{ and }\operatorname{E}[\hat{\boldsymbol{\beta}}_r^*|\boldsymbol{X}_f]=\boldsymbol{0}
		\end{equation}
		with variance:
		\begin{equation}
			\operatorname{Var}[\hat{\boldsymbol{\beta}}_f^*|\boldsymbol{X}_f]= (\sigma^2_Y+\sum_{j \in I_r}\sigma^2_{j}\beta_{j}^2 )(\boldsymbol{X}_f' \boldsymbol{X}_f)^{-1}  \textrm{ and }\operatorname{Var}[\hat{\boldsymbol{\beta}}_r^*|\boldsymbol{X}_f]= \boldsymbol{0} 
		\end{equation}
		We see that the variance is reduced compared to OLS described in equation (\ref{eqOLS})(no correlations and smaller matrix give better conditioning ) for small values of $\sigma_j$ $i.e.$ strong correlations. So we play on the bias-variance tradeoff, reducing the variance by adding a bias. 		
			%There is no theoretical guarantee that our model is better. It's just a compromise between numerical issues caused by correlations for estimation and selection versus increased variability due to structural hypothesis. Therefore we made some simulations to compare both methods (see the end of this paper).
			 \\
	\subsection{Illustration of the tradeoff conveyed by the pretreatment}		  
	 The Mean Squared Error (\textsc{MSE}) on $\hat{\boldsymbol{\beta}}$ is:
	\begin{eqnarray}
		\textsc{MSE}(\hat{\boldsymbol{\beta}}|\boldsymbol{\beta},\boldsymbol{X})&=&\parallel \operatorname{Bias}\parallel_2^2+\operatorname{Tr}(\operatorname{Var}(\hat{\boldsymbol{\beta}})) \\
			&=& 0 + \sigma_Y^2 \operatorname{Tr}((\boldsymbol{X}'\boldsymbol{X})^{-1}) \textrm{ for OLS, and then for the marginal model:} \\
			&=& \parallel\sum_{j \in I_r}\beta_{j}\boldsymbol{\alpha}_j \parallel_2^2 +\parallel \boldsymbol{\beta}_r\parallel^2_2 + (\sigma^2_Y+\sum_{j \in I_2}\sigma^2_{j}\beta_{j}^2 ) \operatorname{Tr}((\boldsymbol{X}_f' \boldsymbol{X}_f)^{-1})
	\end{eqnarray}	 
	To better illustrate the bias-variance tradeoff, we look at a simple example with $p=3$ variables. $\boldsymbol{X}_f$ is composed by two independent scaled Gaussian $\mathcal{N}(0,1)$, $\boldsymbol{x}_3=\boldsymbol{x}_1+\boldsymbol{x}_2+\boldsymbol{\varepsilon}_3$ where $\boldsymbol{\varepsilon}_3\sim{\mathcal{N}(\boldsymbol{0},\sigma_3^2\boldsymbol{I}_n)}$. We also have $\boldsymbol{\beta}=(1,1,1)$ and $\sigma_Y \in \{1,10\}$  . Then we observe the theoretical Mean Squared Error (MSE) of the estimator of both OLS and \textsc{CorReg}'s marginal  model for several values of $\sigma_3$ (strength of the sub-regression) and $n$. Figure \ref{MQE1} shows the theoretical MSE evolution with the strength of the sub-regression:
	\begin{equation}
		1-\mathcal{R}^2=\frac{\operatorname{Var}(\boldsymbol{\varepsilon)_3}}{\operatorname{Var}(\boldsymbol{x}_3)}=\frac{\sigma_3^2}{\sigma_3^2+2}
	\end{equation}

>>>>

\section{Model to select decorrelated covariates}
%\subsection{A classical problem: correlations in regression}\label{sectionOLS}

\subsection{Our proposal: modelisation of the correlations}
We make the hypothesis that $\boldsymbol{X}$ can be described by a partition $\boldsymbol{X}=(\boldsymbol{X}_f,\boldsymbol{X}_r) $ where variables in $\boldsymbol{X}_r$ are endogenous covariates resulting from linear sub-regressions based on $\boldsymbol{X}_f$, the submatrix of mutually independent exogenous covariates. Then $\boldsymbol{X}_r$ is the $n\times p_r$ submatrix of $0\leq p_r <p$ redundent covariates and $\boldsymbol{X}_f$ the $n\times (p-p_r)$ submatrix of the free (independent) covariates.
In the following, we note $\boldsymbol{X}^j$ the $j^{th}$ column of $\boldsymbol{X}$ and $|.|$ represents the cardinal of an ensemble.
The family of $p_r$ regressions within correlated covariates in $\boldsymbol{X}$ is noted:
	\begin{equation}
		\boldsymbol{X}_{r|\boldsymbol{X}_f,S} \textrm{ defined by }\forall \boldsymbol{X}^j \subset \boldsymbol{X}_r: \boldsymbol{X}^j_{|\boldsymbol{X}_f,S}=\boldsymbol{X}_f\boldsymbol{\alpha}_j+\boldsymbol{\varepsilon}_j \textrm{ with } \boldsymbol{\varepsilon}_j \sim\mathcal{N}(\boldsymbol{0},\sigma^2_j\boldsymbol{I}_n) \label{SR}
	\end{equation}
		where $\boldsymbol{\alpha}_j \in \mathcal{R}^{(p-p_r)}$ are the vectors of the regression coefficients between the covariates. So we now have the expression of $P(\boldsymbol{X}_r|\boldsymbol{X}_f) $
\\
\\
We note $I_r$ the set of indices of endogenous variables in $\boldsymbol{X}$ (explained ones).
We also define $I_f=\{I_f^1,\dots,I_f^p \}$ the set of the sets of indices of exogenous covariates (explaining ones $=\boldsymbol{X}_f$) with $\forall j \notin I_r, I_f^j=\emptyset$. 
We see that $I_f$ defines the non-null coefficients in $\boldsymbol{\alpha}_j$ (each sub-regression can be very parsimonious).
Then we have an explicit structure $S=(I_f,I_r,p_f,p_r)$ where $p_r=|I_r|$, $\boldsymbol{p}_f=(p_f^1,\dots,p_f^{p_r})$ is the vector of the number of covariates in each sub-regression  and $p_f^j=|I_f^j|$. 
\\
\\

The partition of $\boldsymbol{X}$ implies the uncrossing rule $I_f\cap I_r=\emptyset$ {\it i.e.} endogenous variables don't explain other covariates. This hypothesis ensures that $S$ contains no cycle and is straightforward readable (no need to order $I_r$). It is not so restrictive because cyclic structures have no sense and any non-cyclic structure can be associated with a structure that verifies the uncrossing constraint by just successively replacing endogenous covariates by their sub-regression when they are also exogenous in some other sub-regressions.

@@

	  We make the choice to distinguish the response variable from the other endogenous variables (that are on the left of a sub-regression). Thus we have one regression on our response variable ($P(\boldsymbol{Y}|\boldsymbol{X}))$ and a system of sub-regressions (without the response variable: $P(\boldsymbol{X}_r|\boldsymbol{X}_f,S)$). Then we consider correlations between the explicative covariates of the main regression, not between the residuals. We see that the $S$ does not depend on $\boldsymbol{Y}$ so it can be learnt independently, even with a larger dataset (if missing values in $\boldsymbol{Y}$).

@@

\subsection{Strategy of use: pretreatment before classical estimation/selection methods}\label{interpretation}
Now we have an explicit structure of correlations and we know which covariates will cause ill-conditioning. The explicit form also gives the ability to understand phenomenon inside the dataset and to see why some covariates are not selected by methods like the LASSO when all $\boldsymbol{X}$ is provided (redundent variables).

As a pretreatment, it allows usage of any method in a second time to estimate $\boldsymbol{\beta}^*$, even with variable selection methods like LASSO or even stepwise \cite{seber2012linear}.

@@

	For the marginal regression model defined in (\ref{modexpl})
%	\begin{equation}
%		\boldsymbol{Y}_{|\boldsymbol{X}_f}= \boldsymbol{X}_f\boldsymbol{\beta}_f^*+ \boldsymbol{\varepsilon}_Y^*
%	\end{equation}			
%		So 
we have the \textsc{OLS} unbiased estimator of $\boldsymbol{\beta}^*$: 
		\begin{equation}
			\hat{\boldsymbol{\beta}}_f^* = (\boldsymbol{X}_f'\boldsymbol{X}_f)^{-1}\boldsymbol{X}_f'\boldsymbol{Y}  \textrm{ and }\boldsymbol{\hat\beta}_r^* = \boldsymbol{0}
		\end{equation}
		We see in (\ref{Trueexpl}) that it gives an unbiased estimation of $\boldsymbol{Y}$ and $\boldsymbol{\beta^*}$
		but in terms of $\boldsymbol{\beta}$ this estimator is biased:
		\begin{equation}
			\operatorname{E}[\hat{\boldsymbol{\beta}}_f^*|\boldsymbol{X}_f]=\boldsymbol{\beta}_f+\sum_{j \in I_r}\beta_{j}\boldsymbol{\alpha}_j \textrm{ and }\operatorname{E}[\hat{\boldsymbol{\beta}}_r^*|\boldsymbol{X}_f]=\boldsymbol{0}
		\end{equation}
		with variance:
		\begin{equation}
			\operatorname{Var}[\hat{\boldsymbol{\beta}}_f^*|\boldsymbol{X}_f]= (\sigma^2_Y+\sum_{j \in I_r}\sigma^2_{j}\beta_{j}^2 )(\boldsymbol{X}_f' \boldsymbol{X}_f)^{-1}  \textrm{ and }\operatorname{Var}[\hat{\boldsymbol{\beta}}_r^*|\boldsymbol{X}_f]= \boldsymbol{0} 
		\end{equation}
		We see that the variance is reduced compared to OLS described in equation (\ref{eqOLS})(no correlations and smaller matrix give better conditioning ) for small values of $\sigma_j$ $i.e.$ strong correlations. So we play on the bias-variance tradeoff, reducing the variance by adding a bias. 		
			%There is no theoretical guarantee that our model is better. It's just a compromise between numerical issues caused by correlations for estimation and selection versus increased variability due to structural hypothesis. Therefore we made some simulations to compare both methods (see the end of this paper).
			 \\
	\subsection{Illustration of the tradeoff conveyed by the pretreatment}	
	Here we look at the OLS case.
For a model defined by (\ref{regressionsimple}), we have the following Ordinary Least Squares (\textsc{OLS}) estimator:
	\begin{equation}
		\boldsymbol{\hat{\beta}}=\left(\boldsymbol{X}'\boldsymbol{X} \right) ^{-1}\boldsymbol{X}'\boldsymbol{Y}
	\end{equation}
	with variance matrix
	\begin{equation}
		\operatorname{Var}(\hat{\boldsymbol{\beta}}_{OLS})=\sigma_Y^2\left(\boldsymbol{X}'\boldsymbol{X} \right) ^{-1} \label{eqOLS}
	\end{equation}
	and without any bias.
	When correlations between covariates are strong, the matrix to invert is ill-conditioned and the variance increases, giving unstable and unusable estimator \cite{hoerl1970ridge}.
	Another problem is that matrix inversion requires $n\geq p$.

	 The Mean Squared Error (\textsc{MSE}) on $\hat{\boldsymbol{\beta}}$ is:
	\begin{eqnarray}
		\textsc{MSE}(\hat{\boldsymbol{\beta}}|\boldsymbol{X})&=&\parallel \operatorname{Bias}\parallel_2^2+\operatorname{Tr}(\operatorname{Var}(\hat{\boldsymbol{\beta}})) \\
			\textsc{MSE}(\hat{\boldsymbol{\beta}}_{OLS}|\boldsymbol{X})&=& 0 + \sigma_Y^2 \operatorname{Tr}((\boldsymbol{X}'\boldsymbol{X})^{-1}) \textrm{ for OLS, and then for the marginal model:} \\
			\textsc{MSE}(\hat{\boldsymbol{\beta}}^*_{OLS}|\boldsymbol{X})&=& \parallel\sum_{j \in I_r}\beta_{j}\boldsymbol{\alpha}_j \parallel_2^2 +\parallel \boldsymbol{\beta}_r\parallel^2_2 + (\sigma^2_Y+\sum_{j \in I_2}\sigma^2_{j}\beta_{j}^2 ) \operatorname{Tr}((\boldsymbol{X}_f' \boldsymbol{X}_f)^{-1})
	\end{eqnarray}	 
	To better illustrate the bias-variance tradeoff, we look at a simple example with $p=3$ variables. $\boldsymbol{X}_f$ is composed by two independent scaled Gaussian $\mathcal{N}(0,1)$, $\boldsymbol{x}_3=\boldsymbol{x}_1+\boldsymbol{x}_2+\boldsymbol{\varepsilon}_3$ where $\boldsymbol{\varepsilon}_3\sim{\mathcal{N}(\boldsymbol{0},\sigma_3^2\boldsymbol{I}_n)}$. We also have $\boldsymbol{\beta}=(1,1,1)$ and $\sigma_Y \in \{1,10\}$  . Then we observe the theoretical Mean Squared Error (MSE) of the estimator of both OLS and \textsc{CorReg}'s marginal  model for several values of $\sigma_3$ (strength of the sub-regression) and $n$. Figure \ref{MQE1} shows the theoretical MSE evolution with the strength of the sub-regression:
	\begin{equation}
		1-\mathcal{R}^2=\frac{\operatorname{Var}(\boldsymbol{\varepsilon)_3}}{\operatorname{Var}(\boldsymbol{x}_3)}=\frac{\sigma_3^2}{\sigma_3^2+2}
	\end{equation}
